Comparison of Terrain Following and Cut Cell Grids using a Non-Hydrostatic Model

### 2016-01-19 src/tf-cutcell-comparison/tf-cutcell-comparison.tex
update resting atmosphere discussion

The initial potential temperature field has $\theta(z = 0) = \SI{288}{\kelvin}$ and a constant static stability with Brunt-V\"ais\"al\"a frequency $N = \SI{0.01}{\per\second}$ everywhere, except for a more stable layer of $N = \SI{0.02}{\per\second}$.  In order to challenge pressure gradient calculations on the cut cell grid, this inversion layer is placed between $\SI{0.5}{\kilo\meter} \leq z \leq \SI{1.5}{\kilo\meter}$ so that it intersects the terrain, unlike \citet{klemp2011} who position the inversion layer above the mountain.
The Exner function of pressure is calculated so that it is in discrete hydrostatic balance in the vertical direction \citep{weller-shahrokhi2014}.  The damping function, \(\mu\), is set to \SI{0}{\per\second}.  Unlike \citet{klemp2011}, there is no eddy diffusion in the equation set.

The test was integrated forward by 5 hours using a timestep $\Delta t = \SI{100}{\second}$ on the BTF, SLEVE and cut cell grids, and a regular grid with flat terrain.
Maximum vertical velocities are presented in figure~\ref{fig:resting} and are similar on the BTF, SLEVE and cut cell grids.  The largest vertical velocity of \SI{0.37}{\meter\per\second} was found on the BTF grid, compared with a maximum of \(\sim \SI{7}{\meter\per\second}\) found by \citet{klemp2011} using their improved horizontal pressure gradient formulation.
Unlike the result from \citet{klemp2011}, the SLEVE grid does not significantly reduce vertical velocities compared to the BTF grid.

@@

The results in figure~\ref{fig:resting} have maximum errors that are comparable with \citet{weller-shahrokhi2014} but, due to the more stable split into implicitly and explicitly treated terms (described in the appendix), the errors decay over time due to the dissipative nature of the advection scheme.

In summary, spurious velocities in the resting atmosphere test were similarly small on all irregular grids, with the cut cell grid offering no improvement over terrain following grids.

@@

\begin{figure}
	\centering
	\includegraphics{../fig-resting/fig-resting.pdf}
%
	\caption{Maximum magnitude of spurious vertical velocity, \(w\) (\si{\meter\per\second}), in the resting atmosphere test with results on BTF, SLEVE, cut cell and regular grids using the model from \citet{weller-shahrokhi2014} which includes a curl-free pressure gradient formulation.}
	\label{fig:resting}
\end{figure}

>>>>

Tests were performed with two different stability profiles, both having an initial potential temperature field has $\theta(z = 0) = \SI{288}{\kelvin}$ and a constant static stability with Brunt-V\"ais\"al\"a frequency $N = \SI{0.01}{\per\second}$ everywhere, except for a more stable layer of $N = \SI{0.02}{\per\second}$.
Figure~\ref{fig:resting}a shows where this inversion layer is positioned in the two tests: the `high inversion' test follows \citet{klemp2011}, placing the layer between $\SI{2}{\kilo\meter} \leq z \leq \SI{3}{\kilo\meter}$; the `low inversion' test is designed to challenge the pressure gradient calculations on the cut cell grid by placing the inversion layer between $\SI{0.5}{\kilo\meter} \leq z \leq \SI{1.5}{\kilo\meter}$ so that it intersects the terrain.

The Exner function of pressure is calculated so that it is in discrete hydrostatic balance in the vertical direction \citep{weller-shahrokhi2014}.  The damping function, \(\mu\), is set to \SI{0}{\per\second}.  Unlike \citet{klemp2011}, there is no eddy diffusion in the equation set.

The test was integrated forward by 5 hours using a timestep $\Delta t = \SI{100}{\second}$ on the BTF, SLEVE and cut cell grids.
Maximum vertical velocities are presented in figure~\ref{fig:resting}b and are similar on the BTF, SLEVE and cut cell grids.  For the high inversion test, the largest vertical velocity of \SI{0.37}{\meter\per\second} was found on the BTF grid after \SI{400}{\second}, compared with a maximum of \(\sim \SI{7}{\meter\per\second}\) found by \citet{klemp2011} using their improved horizontal pressure gradient formulation.
Errors are two orders of magnitude smaller on the cut cell grid with vertical velocities of \(\sim \SI{1e-4}{\meter\per\second}\), but this advantage is lost when the inversion layer is lowered to intersect the terrain.
Unlike the result from \citet{klemp2011}, the SLEVE grid does not further reduce vertical velocities compared to the BTF grid.  This implies that the numerics we are using are less sensitive to grid distortions.

@@

The results in figure~\ref{fig:resting}b have maximum errors that are comparable with \citet{weller-shahrokhi2014} but, due to the more stable split into implicitly and explicitly treated terms (described in the appendix), the errors decay over time due to the dissipative nature of the advection scheme.

In summary, we reproduce the result found by \citet{good2014} that cut cells can reduce spurious velocities over orography.  However, in addition, we find that, with the right numerics, these errors can be very small on a BTF grid.  We also find that, if changes in stratification intersect cut cells, spurious velocities on cut cell grids are comparable with those on TF grids.

@@

\begin{figure}
	\centering
	\includegraphics{../fig-resting/fig-resting.pdf}
%
	\caption{Setup and results of a stratified atmosphere initially at rest.  Tests are performed on four grids for two different stability profiles, with panel (a) showing the placement of the inversion layer in the two profiles.
The low inversion is positioned so that it intersects the terrain, shown immediately above the $x$ axis.
In each test, the inversion layer has a Brunt-V\"ais\"al\"a frequency $N = \SI{0.02}{\per\second}$ and $N = \SI{0.01}{\per\second}$ elsewhere.  
Panel (b) shows the maximum magnitude of spurious vertical velocity, \(w\) (\si{\meter\per\second}), with results on BTF, SLEVE, cut cell and regular grids using the model from \citet{weller-shahrokhi2014} which includes a curl-free pressure gradient formulation.
Results for the high inversion test are shown with solid lines, the low inversion test with dashed lines.}
	\label{fig:resting}
\end{figure}
